---
title: "GTN tutorial"
format: html
---

```{r}
#| message: false
library(DropletUtils)
library(scater)
library(scran)
```

## Data

### Data download

```{r}
#| eval: false
download.file("https://zenodo.org/record/3581213/files/genes.tsv", "data/genes.tsv")
download.file("https://zenodo.org/record/3581213/files/barcodes.tsv", "data/barcodes.tsv")
download.file("https://zenodo.org/record/3581213/files/matrix.mtx", "data/matrix.mtx")
```

### Create a SingleCellExperiment object

```{r}
sce <- read10xCounts(samples = "data")
```

### Inspect the SingleCellExperiment object

```{r}
sce
```

## Preprocessing

### Calculate quality control metrics

```{r}
sce <- scuttle::addPerCellQCMetrics(
  x = sce,
  subsets = list(MT = grep(pattern = "^MT-", x = rowData(sce)$Symbol))
)
# colData(sce)
```


### Basic filters

“Include features detected in at least this many cells”: 3
“Include cells where at least this many features are detected”: 200

```{r}
colData(sce)
```

### Visualise QC Metrics

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "detected"
)
```

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "sum"
)
```

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "subsets_MT_percent"
)
```

```{r}
scater::plotColData(
  object = sce,
  x = "sum",
  y = "detected"
)
```

```{r}
scater::plotColData(
  object = sce,
  x = "sum",
  y = "subsets_MT_percent"
)
```

### Filter Out Low Quality Cells

“Minimum nFeature_RNA”: 200
“Maximum nFeature_RNA”: 2500
“Maximum percent.mt”: 5.0
“Filter by a different metric”: No

```{r}
sce <- sce[, sce$detected >= 200 & sce$detected <= 2500 & sce$subsets_MT_percent <= 5]
sce
```

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "detected"
)
```

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "sum"
)
```

```{r}
#| fig-width: 3
#| fig-height: 5
scater::plotColData(
  object = sce,
  y = "subsets_MT_percent"
)
```

```{r}
scater::plotColData(
  object = sce,
  x = "sum",
  y = "detected"
)
```

```{r}
scater::plotColData(
  object = sce,
  x = "sum",
  y = "subsets_MT_percent"
)
```

## Further Preprocessing

### Normalisation

```{r}
sce <- scuttle::logNormCounts(x = sce)
sce
```

### Identify highly variable genes

```{r}
dec <- scran::modelGeneVar(sce)
hvgs <- scran::getTopHVGs(dec, n = 2000)
```

